---
title: "MVTBoost Example 2: Well-being"
author: "Patrick Miller"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
    fig_caption: yes
vignette: >
  %\VignetteIndexEntry{MVTBoost Example 2: Well-being}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

The purpose of this document is to reproduce the analysis of psychological well-being in (Miller, Lubke, McArtor, & Bergeman, 2015). The reader is referred to the text for details.

# Fitting the model

This scales predictors, and gives them pretty labels.




```{r,eval=TRUE}
#install.packages("mvtboost")
library(mvtboost)
data(wellbeing)

Y <- wellbeing[,21:26]
X <- wellbeing[,1:20]
Ys <- scale(Y)
ynames <- c("Autonomy","Environmental Mastery","Personal Growth","Positive Relationships","Purpose in Life","Self Acceptance")
xnames <- c("Gender","Age","Income","Education","Chronic Health","Somatic Health","Self Report Health","Positive Affect","Negative Affect","Perceived Social Control","Control Internal States","Commitment","Control","Challenge","Ego Resilience","Social Support - Friends","Social Support - Family","Stress-Problems","Stress-Emotions","Loneliness")
cont.id <- unlist(lapply(X,is.numeric))
Xs <- X
Xs[,cont.id] <- scale(X[,cont.id])
colnames(Xs) <- xnames
colnames(Ys) <- ynames
res <- mvtb(Y=Ys,X=Xs)
```

# Tuning the model by 5-Fold CV
```{r,echo=FALSE, eval=FALSE}
# tuning the model
system.time(res5 <- mvtb(Y=Ys,X=Xs,n.trees=10000,shrinkage=.005,cv.folds=5,compress=FALSE))
save(res5,file="wb_cv5.Rdata")
```

```{r,eval=FALSE}
res5 <- mvtb(Y=Ys,X=Xs,n.trees=10000,shrinkage=.005,cv.folds=5,compress=FALSE)
```

## Computing the best number of trees 

```{r,eval=FALSE}
res5$best.trees
summary(res5,covex=FALSE)
```

```{r,echo=FALSE}
load("wb_cv5.Rdata")
res5$best.trees
summary(res5,covex=FALSE)
```



## Showing training and cv error

```{r,echo=FALSE}
plot(x=1:10000,y=res5$trainerr,type="l",ylab="Error",xlab="Number of trees")
abline(v=res5$best.trees$best.cv)
lines(x=1:10000,y=res5$cv.err,type="l",col="red")
legend("topright",legend=c("Training Error","Cross-Validation Error"),lty=c(1,1),col=c("black","red"),bty="n")
```

# Fit

```{r}
yhat <- predict(res5,newdata=Xs)
diag(var(yhat)/var(Ys))
```

# Interpret the model

## Relative influence

```{r,fig.height=5,fig.width=12}
numformat <- function(val){sub("^(-?)0.", "\\1.", sprintf("%.1f", val))}
  
par(mar=c(8,10,1,1))
mvtb.heat(t(mvtb.ri(res5)),clust.method = NULL,cexRow=1,cexCol=1,numformat=numformat)
```

## Covariance explained 
```{r,fig.height=5,fig.width=12}
par(mar=c(8,16.5,1,1),mfrow=c(1,1))
mvtb.heat(res5$covex[,-c(1:7)],cexRow=.9,clust.method = NULL)

```

# Non-linear effects

## Plots

```{r,fig.height=5,fig.width=12}
par(mfcol=c(1,2),mar=c(5,5,4,1))
plot(res5,predictor.no=11,response.no=3,ylim=c(-1,1.5)) # persgrwth on cis
text(-4,1.825,labels="A",xpd=TRUE)
mvtb.perspec(res5,predictor.no=c(11,18),response.no=6)
text(-.5,.5,labels="B",xpd=TRUE)
```

## Departures from addivity

```{r,echo=FALSE,eval=FALSE}
res.nl <- mvtb.nonlin(res5,Y=Ys,X=Xs)
save(res.nl,file="wb_nonlin.Rdata")
```

```{r,eval=FALSE}
res.nl <- mvtb.nonlin(res5,Y=Ys,X=Xs)
```

```{r,echo=FALSE}
load(file="wb_nonlin.Rdata")
```



```{r}
lapply(res.nl,function(r){head(r[[1]])})
```